Arrows for Parallel Computations

### 2017-07-06 paper/content/conclusion.tex
minor. added notes talk with oleg

In order to fix the issues of the Par Monad backend with the |torus| and |ring| skeleton as explained in sections \ref{sec:ring} and \ref{sec:torus} - or any looping skeleton for that matter - additional work that fixes the strictness problems of the |ArrowParallel| instance is required.
\olcomment{Parrows + accelerate = love? Metion port to Frege.}

@@

We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
Of course, we are working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
It would also be interesting to see a hybrid of PArrows and Accelerate.

>>>>

The strictness problems of the |Par| Monad backend materializes with |ArrowLoop|. Additional work is required to ensure correct behaviour of the |Par| Monad in this context.

\mbcomment{mention ArrowLoop in Torus and Ring chapters}
\olcomment{Parrows + accelerate = love? Metion port to Frege.}

@@

We are looking into more experiences with seamless porting of parallel PArrow-based programs across the backends.
Of course, we are working on expanding both our skeleton library and the number of skeleton-based parallel programs that use our DSL to be portable across flavours of parallel Haskells.
It would also be interesting to see a hybrid of PArrows and Accelerate.
Ports of our approach to other languages like Frege or Java directly are in an early development stage.

### 2017-07-06 paper/content/relwork.tex
minor. added notes talk with oleg

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|. \citet{Hughes2005} presents a tutorial on Arrows. Some theoretical details on Arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.

\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\olcomment{ugh, take care!}
\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They basically use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

>>>>

\paragraph{Arrows.}
Arrows were introduced by \citet{HughesArrows}, basically they are a generalised function arrow~|->|. \citet{Hughes2005} presents a tutorial on Arrows. Some theoretical details on Arrows \cite{jacobs_heunen_hasuo_2009,LINDLEY201197,ATKEY201119} are viable. \citet{Paterson:2001:NNA:507669.507664} introduced a new notation for Arrows. Arrows have applications in information flow research \cite{1648705,LI20101974,Russo:2008:LLI:1411286.1411289}, invertible programming \cite{Alimarine:2005:BAA:1088348.1088357}, and quantum computer simulation \cite{vizzotto_altenkirch_sabry_2006}. But probably most prominent application of Arrows is Arrow-based functional reactive programming, AFRP \cite{Nilsson:2002:FRP:581690.581695,Hudak2003,Czaplicki:2013:AFR:2499370.2462161}.
\mbcomment{reference not found!!!}
\citet{Liu:2009:CCA:1631687.1596559} formally define a more special kind of Arrows that capsule the computation more than regular arrows do and thus enable optimizations. Their approach would allow parallel composition, as their special Arrows would not interfere with each other in concurrent execution. In contrast, we capture a whole parallel computation as a single entity: our main instantiation function |parEvalN| makes a single (parallel) Arrow out of list of Arrows.\citet{Huang2007} utilise Arrows for parallelism, but strikingly different from our approach. They basically use Arrows to orchestrate several tasks in robotics. We, however, propose a general interface for parallel programming, while remaining completely in Haskell.

### 2017-07-06 paper/content/topologySkeletons.tex
minor. added notes talk with oleg

Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons. Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}

@@

\subsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

@@

ptorus :: (Arrow arr, Future fut a, Future fut b) =>
	arr (c, a, b) (d, a, b) -> arr (c, fut a, fut b) (d, fut a, fut b)
ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
\end{code} % $
\caption{Definition of the |torus| skeleton. The definitions of |lazyzip3|, |uncurry3| and |threetotwo| have been omitted and can be found in Fig.~\ref{fig:lazyzip3etc}}.
\label{fig:torus}
\end{figure}
As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}). Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
\begin{figure}[tb]
\begin{code}
nodefunction :: Int -> ((Matrix, Matrix), [Matrix], [Matrix]) -> ([Matrix], [Matrix], [Matrix])
nodefunction n ((bA, bB), rows, cols) = ([bSum], bA:nextAs , bB:nextBs)
		where bSum = foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
			     nextAs = take (n-1) rows
			     nextBs = take (n-1) cols
\end{code}
\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}.}
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable. We discuss further examples on larger clusters and in a more detail in the next section.
\begin{figure}[tb]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix Multiplication with a torus (Parrows)]{Matrix Multiplication with a torus (PArrows).}
	\label{fig:torus_parrows_trace}
\end{figure}

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
	\caption[Matrix Multiplication with a torus (Eden)]{Matrix Multiplication with a torus (Eden).}
	\label{fig:torus_eden_trace}
\end{figure}

>>>>

Note that the Par Monad version of the |ArrowParallel| version  presented in this paper (Fig.~\ref{fig:ArrowParallelParMonad}) does not work with the looping behaviour in the |ring| skeletons. Whenever |runPar| is invoked, strictness is enforced, which means that the |lazy| operation in the definition of |ring| will not work.
%Further details on this can be found in \cite{eden_cefp}.

@@

\subsection{Torus skeleton}\label{sec:torus}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the |torus| skeleton.}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a |ring| from Section~\ref{sec:ring} one dimension further, we obtain a |torus| skeleton (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
With our Parallel Arrows we re-implement the |torus| combinator\footnote{Available on Hackage: \url{https://hackage.haskell.org/package/edenskel-2.1.0.0/docs/Control-Parallel-Eden-Topology.html}.} from Eden---yet again with the help of the |ArrowLoop| typeclass.

@@

ptorus :: (Arrow arr, Future fut a, Future fut b) =>
	arr (c, a, b) (d, a, b) -> arr (c, fut a, fut b) (d, fut a, fut b)
ptorus f = arr (\ ~(c, a, b) -> (c, get a, get b)) >>> f >>> arr (\ ~(d, a, b) -> (d, put a, put b))
\end{code} % $
\caption{Definition of the |torus| skeleton. The definitions of |lazyzip3|, |uncurry3| and |threetotwo| have been omitted and can be found in Fig.~\ref{fig:lazyzip3etc}}.
\label{fig:torus}
\end{figure}
As an example of using this skeleton \citep{Loogen2012} showed the matrix multiplication using the Gentleman algorithm (\citeyear{Gentleman1978}). Their instantiation of the skeleton |nodefunction| can be adapted as shown in Fig.~\ref{fig:torusMatMult}.
\begin{figure}[tb]
\begin{code}
nodefunction :: Int -> ((Matrix, Matrix), [Matrix], [Matrix]) -> ([Matrix], [Matrix], [Matrix])
nodefunction n ((bA, bB), rows, cols) = ([bSum], bA:nextAs , bB:nextBs)
		where bSum = foldl' matAdd (matMult bA bB) (zipWith matMult nextAs nextBs)
			     nextAs = take (n-1) rows
			     nextBs = take (n-1) cols
\end{code}
\caption{Adapted |nodefunction| for matrix multiplication with the |torus| from Fig.~\ref{fig:torus}.}
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable. We discuss further examples on larger clusters and in a more detail in the next section.
\begin{figure}[tb]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix Multiplication with |torus| (Parrows)]{Matrix Multiplication with |torus| (PArrows).}
	\label{fig:torus_parrows_trace}
\end{figure}

\begin{figure}[tb]
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_eden_scale}
	\caption[Matrix Multiplication with |torus| (Eden)]{Matrix Multiplication with |torus| (Eden).}
	\label{fig:torus_eden_trace}
\end{figure}
